---
title: 'Calculating a time series of diversification rates in a time-calibrated molecular
  phylogeny '
output:
  html_document:
    toc: yes
  pdf_document:
    fig_caption: yes
    highlight: tango
    toc: yes
---




I want to describe the calculation of timeseries of diversification rates that I used in Simpson et al. (2011) and Simpson (2013). 

The basic method is simple and there is a function for calculating the rate estimate and likelihood in a specified window of time in the laser package by Rabosky: `yule.Window()`. But there are a couple of tricks to using it effectively, especially if you want to use an arbitary timescale. 

```{r yulewindow, cache=TRUE}
# This is yuleWindow from laser. But importantly, I've commented out the text
# outputs. This speeds things up and makes this document shorter.

yuleWindow <- function (x, st1 = x[1], st2 = 0) 
{
    checkbasal(x)
    if (!is.numeric(x)) 
        stop("object x not of class 'numeric'")
    if (st1 <= st2) {
        cat("Error: invalid input\n\n")
        cat("shift points must be in temporal order (first, second)\n")
        cat("and in units of time/substitutions before present\n\n")
        return()
    }
    x <- rev(sort(x))
    res <- as.list(yuleint2(x, st1, st2))
    if (is.na(res$LH)) {
        print("Error.  Must be >= 1 speciation event within temporal window\n")
        stop()
    }
   # cat("---------------------------------\n")
    #cat("Results of fitting Yule model to temporal window of \n")
    #cat("(", st1, ",", st2, ") time units before present:\n\n")
    #cat("ML speciation rate: ", res$smax, "\n\nLikelihood: ", 
    #    res$LH, "\n\n")
    return(res)
}
```


Tanja Stadler published a similar method (Stadler 2011 and the package `TreePar`) that allows rates to change from interval to interval only if the likelihood in that interval exceeds a certain threshold defined by a constant rate. We know from the fossil record that rates are never constant, so having a null that is constant just does not make sense to me. However, you can easily test for that with the code here by saving the likelihoods.


For a single simulated tree: 
```{r sim tree, cache=TRUE}
library(ape)

b <- 0.2
d <- 0.1
C <- 10
R <- 1
tree <-rbdtree(b, d, Tmax = 50)
plot(tree, show.tip.label = F)
```


You don't even need a real time scale. In the example below, the scale is simply a window 1.5 million years wide sliding every 1 million years. Now with this kind of timescale there are two options that become important. 1) the width of the window (`w`), and 2)the minimum number of nodes in the window you want to calculate rates on (`n`). These two paramters interact with each other, the wider the window the easier it is to get lots of nodes. All of this is not important if you use a geological time scale and multiple trees pulled from the posterior of a Bayesian analysis.


The basic code is with an arbitrary timescale is:
```{r single tree, cache=TRUE}
suppressMessages(library(laser))

max.t <- max(branching.times(tree))


t <- max.t:1
div <- as.numeric()
lik <- as.numeric()

w=1.5 
# w is the width of the window. This can be adjusted. Wider windows will
# natrually include more nodes and give a better estimate of the diversification
# rates.

n=1   
# n is the minimum number of nodes needed in a window. This needs to be larger
# than 0, but as small as possible



  bt.tree<-branching.times(tree)

for(i in t){
  a <- length(which(bt.tree > t[i] & bt.tree < t[i]+w)  )
  if(a > n){
    yw <- try(yuleWindow(bt.tree, st1=t[i]+w, st2=t[i]))
    div[i] <- yw$smax
    lik[i] <- yw$LH
    
    } else{div[i]<-NA
           lik[i] <- NA}
}
```

The diversification rates are saved in `div` and the likelihoods are in `lik`.
```{r div rate plot}
plot(t, div, type = 'l', xlim = c(50,0), xlab = "Time (Ma)", ylab = "Diversification Rate")
```

```{r likelihood plot}
plot(t, lik, type = 'l', xlim = c(50,0), xlab = "Time (Ma)", ylab = "Likelihood")
```




Adding a geological timescale
---

This is how I prefer to do things.

First I define the time scale. Usually I keep a csv file with the timescale and source it.
```{r timescale, cache=TRUE}
#Cenozoic timescale
tscale.bot <- c(66, 61.6,59.2, 56, 47.8,
                41.2,37.8, 33.9, 
                28.1,23.0, 20.4, 
                16.0, 13.8, 11.6, 
                7.2, 5.3, 3.6, 2.6, 1.8, 0.01)
tscale.top <- c(tscale.bot[-1], 0)

tscale <- cbind(tscale.bot, tscale.top, tscale.top + (tscale.bot-tscale.top)/2)
tscale

```

Then I run the rate calculation.
```{r rate calc, cache=TRUE}
suppressMessages(library(laser))
div <- as.numeric()

# n=0   # n is the minimum number of nodes needed in a window. 
# This needs to be larger than 0, but as small as possible

  bt.tree<-branching.times(tree)

  for(i in 1:length(tscale[,1])){
    a <- length(which(bt.tree > tscale[i,2] & bt.tree < tscale[i,1]))
    if(a > n){
      div[i] <- try(yuleWindow(bt.tree, st1=tscale[i,1], st2=tscale[i,2])$smax)
    } else{div[i]<-NA
           }
  }
```

After running the calculation, I plot it...
```{r plot calc, fig.width=12}
plot(tscale[,3], div, 
     type = 'n', 
     xlim = c(66,0), 
     ylim = c(0, 0.3), 
     xlab = "", ylab = "", 
     axes = F)
# draw timescale in background

bases <- c(66, 56, 33.9, 23.0, 5.3, 2.6, 0.01,0.01)
cc<-rep(c("grey85","grey90"),11) 

bt <- 0.05; to <- 0.3
for (b in 1:16) rect(xleft=bases[b], ybottom=bt, 
                     xright=bases[b+1], ytop=to,col=cc[b], border=NA)

# add ribbon of time scale
 bt <- 0.03; to <- 0.05
 # Add boxes  for legend of time 
  for (b in 1:6) rect(xleft=bases[b], ybottom=bt, 
                      xright=bases[b+1], ytop=to,col=NA,border='grey50')

# Add Text for legend of time
    bt <- (bt+to)/2 # Level of text
    tpl <- c(61.6, 45, 28.1, 14, 4, 1.1)
    tx <- c( "Paleocene", "Eocene", "Oligocene", "Miocene", "Pli", "Ple")

for (t in 1:6) text(x=tpl[t], y=bt, labels=tx[t])

axis(1, col = "grey75", line = -1.5)
mtext("Time (Ma)", side = 1, line = 0.5)
axis(2, col = "grey75", line = 0, at = c(0.05, 0.1, 0.15, 0.2, 0.25, 0.3))
mtext("Diversification Rate", side = 2, line = 2.5)

# plot lines
lines(tscale[,3], div)
```

In the background of this plot I show epoch-level geological time. Also, I usually use a function for plotting these timescales in the background.




References
---
Rabosky, Daniel L. "LASER: a maximum likelihood toolkit for detecting temporal shifts in diversification rates from molecular phylogenies." Evolutionary Bioinformatics Online 2 (2006): 247.

Simpson, Carl, et al. "Evolutionary diversification of reef corals: a comparison of the molecular and fossil records." Evolution 65.11 (2011): 3274-3284.

Simpson, Carl. "Species selection and the macroevolution of coral coloniality and photosymbiosis." Evolution 67.6 (2013): 1607-1621.

Stadler, Tanja. "Mammalian phylogeny reveals recent diversification rate shifts." Proceedings of the National Academy of Sciences 108.15 (2011): 6187-6192.